---
title: "score_comparisons"
author: "Yesol Sapozhnikov"
date: "8/18/2020"
output: html_document
---

This document compares scores from my assay data to others (e.g. SDM viability data, Zoe's data, etc). For this analysis, consider t70_repX for now as before.


Packages and data import
```{r}
library(tidyverse)
viability <- read.csv("~/Projects/phix_host/phix_jt/Craigs_21Site_ddG_Viability_data.csv")[,-c(1:2)] # SDM viability data
yesol_df2 <- read.csv("../data/dada2result_yesol2.csv", stringsAsFactors = F)[,-1] # Vars count data
combined_scores <- read.csv("../data/combined_var_scores_yesol.csv") # scores prev. calculated
# Assign values to NAs:
yesol_df2$AAsub[is.na(yesol_df2$AAsub)] <- "wt"
yesol_df2$nmut_offtarget[is.na(yesol_df2$nmut_offtarget)] <- 0
```



Create a new dataframe of each variant's counts over time points 
```{r}
vars <- unique(yesol_df2$AAsub)
all_vars_wd <- as.data.frame(vars)
samples <- unique(yesol_df2$sample)
for (i in 1:length(samples)){
  df_i <- filter(yesol_df2, sample == samples[i])
  for (j in 1:length(vars)){
    if (vars[j] %in% df_i$AAsub){
      all_vars_wd[j, samples[i]] <- df_i[df_i$AAsub==vars[j],]$count
    } else {
      all_vars_wd[j, samples[i]] <- 0
    }
  }
}
```

Filter out variants that appear only in ligation mix and score separately.
```{r}
# Subset out rows of variants only present in lig mix
ligs_idx <- which(rowSums(all_vars_wd[ , 5:17]) == 0)
ligs_idx <- c(1, ligs_idx) # add first row (wt)
ligs_wd <- all_vars_wd[ligs_idx, ] %>% 
  select(vars, starts_with("lig"), t35_rep1:t35_rep3, t70_rep1:t70_rep3)
# change colname "lig_repX" to "t0_repX" 
colnames(ligs_wd)[2:4] <- c("t0_rep1","t0_rep2","t0_rep3")
# Gather counts into a long form, create additional columns for scoring
ligs <- gather(ligs_wd, t0_rep1:t70_rep3, key = "sample", value = "count")
for (i in 1:nrow(ligs)){
  ligs$time[i] <- as.numeric(sub("t","",unlist(str_split(ligs$sample[i], "_r"))[1]))
  ligs$rep[i] <- unlist(str_split(ligs$sample[i], "_"))[2]
}

# Score
# Following function will generate a list of 3 dataframes: scores, se, p-values.
source("../scripts/fn_calc_score.R")
ligs_scores <- calc_score(ligs, c(0, 35, 70), 3)

vars <- ligs_scores$scores$vars # get a vector of var names
ligs_combined_scores <- as.data.frame(matrix(data = NA, nrow = length(vars), ncol = 4))
colnames(ligs_combined_scores) <- c("vars","betaML","var_betaML","eps")
ligs_combined_scores$vars <- vars

# Following function will combine replicate scores and output a list of:
# combined score, variance, and eps
source("../scripts/fn_calc_rmle.R")
# Run this function row by row (vars):
for (j in 2:length(vars)){ # skip wt
  y <- as.numeric(ligs_scores$scores[j,-1]) # y is a num vector of replicate scores (slope) 
  sigma2i <- (as.numeric(ligs_scores$se[j,-1]))^2 # square of SE
  outlist <- calc_rmle(y, sigma2i, 50) # iteration = 50 in the paper but try smaller number and check eps
  ligs_combined_scores$betaML[j] <- outlist$betaML
  ligs_combined_scores$var_betaML[j] <- outlist$var_betaML
  ligs_combined_scores$eps[j] <- outlist$eps
}
```



## Comparing with SDM viability data

Correlation with viability data
```{r}
# Extract necessasry SDM viability columns and rename
viability <- select(viability, mutation, obs)
colnames(viability) <- c("vars","obs")
# Combine two score dfs (one previously done, one that are from lig mix as t0)
scores <- bind_rows(combined_scores[ , 1:3], ligs_combined_scores[ , 1:3]) 
# Join with viability data
joined <- inner_join(viability, scores, by = "vars")
# Regress binary viability on fitness score
fit <- glm(obs~betaML, data = joined, family = binomial)
```
Plot
```{r}
x <- seq(-.2, 0.11, 0.01)
y <- predict(fit, list(betaML = x), type = "response")
plot(joined$obs~joined$betaML, pch = 20, main = "Fitness score vs Viability",
     xlab = "Score (beta MLE, Enrich2 algorithm)", ylab = "Viability")
lines(x, y, col = "blue")
```
Try weighted regression
```{r}
# add weight column (1/var)
joined$weight <- 1/joined$var_betaML
fit_weighted <- glm(obs~betaML, data = joined, weights = weight, family = binomial)
```
Plot
```{r}
x <- seq(-.2, 0.11, 0.01)
y <- predict(fit_weighted, list(betaML = x), type = "response")
plot(joined$obs~joined$betaML, pch = 20, main = "Fitness score (weighted) vs Viability",
     xlab = "Score (beta MLE, Enrich2 algorithm)", ylab = "Viability")
lines(x, y, col = "blue")
```

```{r}
ggplot(joined[-19,], aes(x = betaML, y = obs, size = weight)) +
  geom_point()
```

Without ligs-only variants, but do weighted regression
```{r}
# Join with viability data
joined2 <- inner_join(viability, combined_scores[ , 1:3], by = "vars")
# Regress binary viability on fitness score
fit2 <- glm(obs~betaML, data = joined, family = binomial)
# plot
x <- seq(-.2, 0.11, 0.01)
y <- predict(fit2, list(betaML = x), type = "response")
plot(joined2$obs~joined2$betaML, pch = 20, main = "Fitness score vs Viability",
     xlab = "Score (beta MLE, Enrich2 algorithm)", ylab = "Viability")
lines(x, y, col = "blue")
```
Weighted:
```{r}
joined2$weight <- 1/joined2$var_betaML
fit2_weighted <- glm(obs~betaML, data = joined2, weights = weight, family = binomial)
x <- seq(-.2, 0.11, 0.01)
y <- predict(fit2_weighted, list(betaML = x), type = "response")
plot(joined2$obs~joined2$betaML, pch = 20, main = "Fitness score (weighted) vs Viability",
     xlab = "Score (beta MLE, Enrich2 algorithm)", ylab = "Viability")
lines(x, y, col = "blue")
```



## Distribution of Fitness Effects

Histogram of fitness for whole data (minus lig-only variants)
```{r}
# Create a column to indicate variant types (single, double, etc)
for (i in 1:nrow(combined_scores)){
  nmut_i <- length(unlist(str_split(combined_scores$vars[i], ",")))
  if (nmut_i == 1){
    combined_scores$type[i] <- "single"
  } else if (nmut_i == 2){
    combined_scores$type[i] <- "double"
  } else {
    combined_scores$type[i] <- "more"
  }
}

# histogram of betaML grouped by type
ggplot(combined_scores, aes(x = betaML)) + 
  geom_histogram() +
  facet_wrap(~type)
```

Histogram of fitness for variants that have SDM viability data
```{r}
joined2$obs <- as.factor(joined2$obs)
ggplot(joined2, aes(x = betaML, fill = obs)) +
  geom_histogram() +
  scale_fill_manual(values=c("black", "blue"))
```

What causes the bimodal shape, especially the large counts near ~ -0.10? Check if majority of the left peak is due to lethal variants.
```{r}
var_counts <- select(all_vars_wd, vars, t0:t35_rep3, t70_rep1:t70_rep3)
var_counts <- var_counts[-ligs_idx, ] # remove lig-only vars (ligs_idx includes wt too)
# Get lethals (vars present only in t0)
lethals_idx <- which(rowSums(var_counts[, 3:8]) == 0)
lethal_vars <- var_counts[lethals_idx, "vars"] # a vector of var names that are lethal
# Get 3rd quartile of SE
lg_se <- summary(combined_scores$se_betaML)["3rd Qu."]
# Create a new "attributes" column to indicate misc info such as lethals, large SE, etc
var_scores <- combined_scores[-1,] # Get rid of the wt row and rename the df
for (i in 1:nrow(var_scores)){
  if (var_scores$vars[i] %in% lethal_vars){
    var_scores$attr[i] <- "lethal"
  } else if (var_scores$se_betaML[i] > lg_se){
    var_scores$attr[i] <- "large_se" # some lethals will fall in this category but lehtla attr more imp
  } else {var_scores$attr[i] <- "none"}
}

ggplot(var_scores, aes(x = betaML, color = attr)) + 
  geom_histogram() +
  facet_wrap(~type)
```


## Distribution of Fitness Effects after Removing One-timers (including lethals)

Add obs_sample column to indicate how many samples have non-zero read counts:
```{r}
for (i in 1:nrow(var_counts)){
  j = 0
  for (s in 2:8){ #check sample by sample and add to j whenever non-zero
    if (var_counts[i, s] != 0){
      j = j + 1
    }
  }
  var_counts$obs_samples[i] <- j
}
# create the obs_samples column in the scores df as well
var_scores$obs_samples <- var_counts$obs_samples
```

Plot histograms only with variants that occur in more than one sample:
```{r}
var_scores %>% filter(obs_samples > 1) %>%
  ggplot(aes(x = betaML, color = attr)) +
  geom_histogram() +
  facet_wrap(~type)
```
Histogram of total (without faceting by type)
```{r}
var_scores %>% filter(obs_samples > 1) %>%
  ggplot(aes(x = betaML, color = attr)) +
  geom_histogram()
```


## More on One-timers 

Label which sample the variants are observed (t0? t70_rep2?)
```{r}
samples <- names(var_counts)[2:8]
for (i in 1:nrow(var_counts)){
  if (var_counts$obs_samples[i] == 1) {
    for (j in 1:length(samples)){
      if (var_counts[i, samples[j]] != 0){
        var_counts$obs_in[i] <- samples[j]
      }
    }
  } else {
    var_counts$obs_in[i] <- "mult"
  }
}
# Copy this newly created column to var_score df as well
var_scores$obs_in <- var_counts$obs_in
table(var_counts$obs_in)
```

Score and SE distribution of these one-timers grouped by sample they occur
```{r}
ggplot(var_scores, aes(x=obs_in, y=betaML)) +
  geom_boxplot()
ggplot(var_scores, aes(x=obs_in, y=se_betaML)) +
  geom_boxplot()
```



## More on SDM Viability Data 

Categories to label SDM data:
  - lig (only observed in ligation mix)
  - lethal (only observed in t0)
  - deleterious
  - neutral
  - beneficial
  
How to determine deleterious to beneficial scores? - quartiles?
```{r}
var_scores %>% filter(attr != "lethal") %>%
  summary(var_scores)
```
=> -0.012 < "neutral" < 0.0053


```{r}
# a vector of variant names for lig-only
lig <- ligs_wd[-1,]$vars
# lethal variants are already done: lethal_vars

# Create a new column to categorize each var in joined df
for (i in 1:nrow(joined)){
  if (joined$vars[i] %in% lig){
    joined$category[i] <- "lig"
  } else if (joined$vars[i] %in% lethal_vars){
    joined$category[i] <- "lethal"
  } else if (joined$betaML[i] < -0.012){
    joined$category[i] <- "deleterious"
  } else if (joined$betaML[i] > 0.0053){
    joined$category[i] <- "beneficial"
  } else {
    joined$category[i] <- "neutral"
  }
}
```

Get counts of variants for each category:
For inviables:
```{r}
table(filter(joined, obs == 0)$category)
```
For viables:
```{r}
table(filter(joined, obs == 1)$category)
```


## Comparing with Zoe's data ("mix" samples)

Read in csv and score:
```{r}
zoe_df2 <- read.csv("../data/dada2result_zoe2.csv", stringsAsFactors = F)[,-1]
zoe_df2$AAsub[is.na(zoe_df2$AAsub)] <- "wt"
# make wide-form df to get counts by samples as columns
vars_z <- unique(zoe_df2$AAsub)
zoe_vars_wd <- as.data.frame(vars_z)
samples <- unique(zoe_df2$sample)
for (i in 1:length(samples)){
  df_i <- filter(zoe_df2, sample == samples[i])
  for (j in 1:length(vars_z)){
    if (vars_z[j] %in% df_i$AAsub){
      zoe_vars_wd[j, samples[i]] <- df_i[df_i$AAsub==vars_z[j],]$count
    } else {
      zoe_vars_wd[j, samples[i]] <- 0
    }
  }
}

# select "mix" samples only
mix_wd <- select(zoe_vars_wd, vars_z, starts_with("mix"))
# filter out variants with zero counts in all time points in this subsetted dataframe
mix_wd <- mix_wd[rowSums(mix_wd[,-1]) != 0, ]


# Convert to a long form
zoe_vars <- gather(mix_wd, mix_t0_9_27:mix_t70_9_27, key = "sample", value = "count")
# Add additional columns needed for scoring
for (i in 1:nrow(zoe_vars)){
  zoe_vars$time[i] <- as.numeric(sub("t","",unlist(str_split(zoe_vars$sample[i],"_"))[2]))
  zoe_vars$date[i] <- unlist(str_split(zoe_vars$sample[i], "_"))[4]
}
# fix rep column
zoe_vars <- mutate(zoe_vars, 
                   rep = case_when(date == "27" ~ "rep1",
                                   date == "12" ~ "rep2",
                                   date == "24" ~ "rep3",
                                   date == "04" ~ "rep4"))
# Score
source("../scripts/fn_calc_score.R")
zoe_scores <- calc_score(zoe_vars, c(0, 30, 50, 70), 4)
vars <- zoe_scores$scores$vars
zoe_comb_scores <- as.data.frame(matrix(data = NA, nrow = length(vars), ncol = 4))
colnames(zoe_comb_scores) <- c("vars", "betaML","var_betaML","eps")
zoe_comb_scores$vars <- vars
source("../scripts/fn_calc_rmle.R")
for (j in 2:length(vars)){
  y <- as.numeric(zoe_scores$scores[j,-1])
  sigma2i <- (as.numeric(zoe_scores$se[j,-1]))^2
  outlist <- calc_rmle(y, sigma2i, 50)
  zoe_comb_scores$betaML[j] <- outlist$betaML
  zoe_comb_scores$var_betaML[j] <- outlist$var_betaML
  zoe_comb_scores$eps[j] <- outlist$eps
}
```


Get scored variants into a new df for further inspection
```{r}
zvar_scores <- zoe_comb_scores[-1, 1:3] # just get vars, beta, variance columns
zvar_scores$se <- sqrt(zvar_scores$var_betaML)
# Create a column to indicate variant types (single, double, etc)
for (i in 1:nrow(zvar_scores)){
  nmut_i <- length(unlist(str_split(zvar_scores$vars[i], ",")))
  if (nmut_i == 1){
    zvar_scores$type[i] <- "single"
  } else if (nmut_i == 2){
    zvar_scores$type[i] <- "double"
  } else {
    zvar_scores$type[i] <- "more"
  }
}
# Get analogous counts df to use in conjunction
zvar_counts <- mix_wd[-1,] # mix_wd is the counts df with zero variants filtered out. Just remove wt row.

# histogram of betaML grouped by type
ggplot(zvar_scores, aes(x = betaML)) + 
  geom_histogram() +
  facet_wrap(~type)
```


Put in other information for analysis
```{r}
# lethals?
# Get lethals (vars present only in t0)
lethals_idx_z <- which(rowSums(zvar_counts[, 3:14]) == 0)
lethal_vars_z <- zvar_counts[lethals_idx_z, "vars_z"] # a vector of var names that are lethal
# Get 3rd quartile of SE
lg_se <- summary(zvar_scores$se)["3rd Qu."]
# Create a new "attributes" column to indicate misc info such as lethals, large SE, etc
for (i in 1:nrow(zvar_scores)){
  if (zvar_scores$vars[i] %in% lethal_vars_z){
    zvar_scores$attr[i] <- "lethal"
  } else if (zvar_scores$se[i] > lg_se){
    zvar_scores$attr[i] <- "large_se" # some lethals will fall in this category but lehtla attr more imp
  } else {zvar_scores$attr[i] <- "none"}
}
# histograms
ggplot(zvar_scores, aes(x = betaML, color = attr)) + 
  geom_histogram() +
  facet_wrap(~type)
```

Find variants present both mine and Zoe's and compare their fitness score
```{r}
# A vector of common variant names
yz <- intersect(var_scores$vars, zvar_scores$vars)
# Subset out respective df
yz_y <- var_scores[var_scores$vars %in% yz,] %>% arrange(vars)
yz_z <- zvar_scores[zvar_scores$vars %in% yz,] %>% arrange(vars)
common <- data.frame(vars=yz_y$vars, yesol=yz_y$betaML, mix=yz_z$betaML)
ggplot(data = common, aes(x=yesol, y=mix)) +
  geom_point() +
  geom_smooth(method=lm, se=FALSE) +
  annotate(geom="text", label="r=0.22", color="red", x=0.06, y=-0.06)
```

Try using variants with clearly high/low scores
```{r}
# Mix samples quartiles
summary(zvar_scores$betaML)
```
Mix samples cutoffs:
  -"high" score: > 0.0048
  -"low" scores: < -0.011

```{r}
mix_filtered <- filter(zvar_scores, betaML > 0.0048 | betaML < -0.011)
yesol_filtered <- filter(var_scores, betaML > 0.0053 | betaML < -0.012)
yz2 <- intersect(mix_filtered$vars, yesol_filtered$vars)
yz2_y <- yesol_filtered[yesol_filtered$vars %in% yz2,] %>% arrange(vars)
yz2_z <- mix_filtered[mix_filtered$vars %in% yz2,] %>% arrange(vars)
common_filt <- data.frame(vars=yz2_y$vars, yesol=yz2_y$betaML, mix=yz2_z$betaML)
ggplot(data = common_filt, aes(x=yesol, y=mix)) +
  geom_point() +
  geom_smooth(method=lm, se=FALSE) +
  annotate(geom="text", label="r=0.4", color="red", x=0.06, y=-0.06)
```


Check correlations between replicates
```{r}
# Need the intermediate df -- run scoring chunks from post_dada2_analysis.Rmd first
var_counts_filt <- filter(var_counts, obs_samples == 7)
vars_filt <- var_counts_filt$vars
var_repscores_filt <- yesol_scores$scores[yesol_scores$scores$vars %in% vars_filt,] #Need the score list
# correlation test
cor(var_repscores_filt[,2:4])
```

```{r}
plot(var_repscores_filt[,2:4])
```

Try the same with Zoe's mix samples
```{r}
for (i in 1:nrow(zvar_counts)){
  j = 0
  for (s in 2:14){ 
    if (zvar_counts[i, s] != 0){
      j = j + 1
    }
  }
  zvar_counts$obs_samples[i] <- j
}
zvar_counts_filt <- filter(zvar_counts,obs_samples >= 5)
zvars_filt <- zvar_counts_filt$vars_z
zvar_repscores_filt <- zoe_scores$scores[zoe_scores$scores$vars %in% zvars_filt,]

cor(zvar_repscores_filt[,2:5])
```
```{r}
plot(zvar_repscores_filt[,2:5])
```


How consistent are our counts??
```{r}
repscores <- zoe_scores$scores %>% gather(rep1:rep4, key = "reps", value = "score")
se <- zoe_scores$se %>% gather(rep1:rep4, key = "reps", value = "se")
ggplot(data = se, aes(x=se)) +
  geom_histogram() +
  geom_freqpoly(data = repscores, aes(x=score), color = "blue") +
  facet_wrap(~reps)
```
```{r}
repscores2 <- yesol_scores$scores %>% gather(rep1:rep3, key = "reps", value = "score")
se2 <- yesol_scores$se %>% gather(rep1:rep3, key = "reps", value = "se")
ggplot(data = se2, aes(x=se)) +
  geom_histogram() +
  geom_freqpoly(data = repscores2, aes(x=score), color = "blue") +
  facet_wrap(~reps)
```

## Comparing with fitness score

Some G mutants have fitness scores from liquid growth assay by Lu. Plot fitness against beta to see if Enrich2 scoring agrees with traditional fitness assay.

```{r}
fitness <- read.csv("../data/Gmut_fitness_Lu.csv")
muts <- fitness$mutant[-7] # make a vector of mutants except wt
combined_scores %>% filter(vars == "G118I")
```
=> none found.



